---
title: "Intermediate R"
format: html
editor: visual
---

## Prerequisites

You should have a basic working knowledge of R/RStudio, up to the level covered in the **Getting Started with R & RStudio** workshop \[link\].

## Exercises

Exercises...... \[see LMMs\]

## Data

The data used in these materials is the `penguins` data from the palmerpenguins package.

```{r}
#| eval: false
install.packages("palmerpenguins")
```

Load

```{r}
library(palmerpenguins)

penguins <- data.frame(penguins)
```

\^ check I want to do this way

## Exploring the data

## The dplyr package

The `dplyr` package includes various functions for common data manipulation tasks, such as filtering rows, selecting specific columns, sorting, modifying or removing columns, and summarising data.

This kind of data manipulation is possible in base R, but `dplyr` makes data exploration and data analysis easy and fast in R.

If you have not previously installed `dplyr`.............

```{r}
#| eval: false
install.packages("dplyr")
library(dplyr)
```

### Useful functions in dplyr

`dplyr` offers a range of useful functions, each designed to perform a different data manipulation task. Some key ones are grouped below, grouped by the component of the data they work with (rows, columns, or groups of rows): \[\[EDIT\]\]

|   | dplyr function | Description |
|-------------------|-------------------|------------------------------------|
| 3.1 Row operations | `filter(...)` | Filter rows (choose rows based on column values) |
|  | `slice(...)` | Choose rows based on location |
|  | `arrange(...)` | Sort the data (change the order of the rows) |
|  | `desc(...)` |  |
|  | `distinct(...)` |  |
| Column operations | `select(...)` | Select columns to be included in the data |
|  | `rename(...)` | Change the name of columns |
|  | `mutate(...)` | Create new columns |
| Summarisation | `group_by(...)` | Group the data by the values of (a) column(s) |
|  | `summarise(...)` | Summarise (or aggregate the data - collapse a group into a single row |
|  | `ungroup(...)` |  |
| Joins | `left_join(...)` |  |
|  | `right_join(...)` |  |
|  | `inner_join(...)` |  |
|  | `full_join(...)` |  |
|  |  |  |

For a complete list of all dplyr functions, you can check official `dplyr` documentation directly:

-   dplyr Package Reference: <https://dplyr.tidyverse.org/reference/index.html>

-   CRAN Package Page: <https://cran.r-project.org/web/packages/dplyr/>

### 3.0 General syntax 

Do general syntax?

### The pipe operator

An operator in R is a symbol that perform an operation (action) on values or objects.

The `%>%` operator in R is called the pipe operator, and it comes from the `magrittr` package (which is included in `dplyr`). It allows for cleaner and more readable code by avoiding nested function calls or the need to save intermediate objects.

It passes the result of one function as the first argument to the next function; the result from one step is then “piped” into the next step You can use the pipe to rewrite multiple operations that you can read left-to-right, top-to-bottom (reading the pipe operator as “then”)

### 3.1 Row operations

The `filter(...)` function allows you to select a subset of rows in a data frame.

<details>

<summary>When might filtering be useful?</summary>

This text is hidden until expanded.

-   Bullet 1
-   Bullet 2

You can include **Markdown** here.

</details>

The first argument is the name of the data frame you want to work with.

The second and subsequent arguments refer to variables within that data frame, selecting rows where the expression is true. **Write better**

**General syntax:**

```{r}
#| eval: false
filter(data, condition1, condition2, ...)
```

For example, filter for penguins with body mass greater than 4000g. R checks every row of the data frame `penguins` to see whether the condition `body_mass_g > 4000` to.... and return all rows where TRUE (explain).

```{r}
#| eval: false
filter(penguins, body_mass_g > 4000)
```

Further examples:

If you want to check for equality.....

```{r}
#| eval: false
filter(penguins, Species == "Adelie")
filter(penguins, year == "2007")
```

Multiple conditions.........

```{r}
#| eval: false
filter(penguins, Species == "Adelie", body_mass_g > 4000)
```

\[for each one show base R version\]

### 3.2 Column operations

\[Do a dropdown linking to base R data manipulation section, and showing that using xx\$xx \<- is easier than mutate, especially when only doing one variable\]

===========================

\[maybe put this as a general "warning" reminder somwhere - see red in LMMs\]

If you run the above code, the rows will print in the console. Perhaps you want to save this as a new dataset and do some analysis on it, in which case you need to assign it to a new variable name using the assingment operator (`<-`).

```{r}
#| eval: false
heavy_penguins <- filter(penguins, body_mass_g > 4000)
```

## 
