# Workshop 2

## Learning outcomes

::: {.callout-tip icon="false"}
### [{{< fa regular lightbulb >}}]{style="color: #872046;"}  Learning outcomes

By the end of this workshop, you should be able to:

- xxxxxxxxxxxxxx
- xxxxxxxxxxxxxx
:::

## Workshop structure

## Setup {#setup}

Follow the steps below to get set up for this workshop:

1.  Open RStudio (you do not need to open R separately)

2.  Create a new script and save it in a location you can easily find later

3.  To install all of the packages used in this course, run the following code in your **console**:

::: {.panel-tabset .panel-green group="language"}
## R

```{r}
#| eval: false
# Install packages
install.packages("palmerpenguins")
install.packages("dplyr")
install.packages("ggplot2")
```

{{< fa triangle-exclamation >}} If you have previously installed any of these packages on your device, you do not need to install them again.
:::

4.  To *load* the required packages, copy and run the following code in your **script**:

::: {.panel-tabset .panel-green group="language"}
## R

```{r}
#| message: false
#| warning: false
#| results: "hide"
# Load packages
library("palmerpenguins")
library(dplyr)
library(ggplot2)
```

{{< fa lightbulb >}} You'll need to load any packages required for your work each time you start a new R session.
:::

::: {style="margin-top: 2em;"}
:::

::: callout-important
### Write code in your script, not the console

From now on, unless you're testing a small piece of code, write all of your code in the script rather than the console. This makes it easier to save, review, and rerun your work later.
:::

## Reminders


## Data:penguins















