{{< include _intermediate_r_setup.qmd >}}

# Introduction to linear regression

We'll look at an example of simple linear regression.

Assumptions:

- **linearity**: there is a linear (straight-line) relationship between the independent variable (predictor) and dependent variable (outcome)
- **independent**: observations are independent of each other
- **homoscedasticity**: residual errors have a constant variance across all levels of the independent variable
- **normality**: the residual errors are approximately normally distributed

## Data

::: {.panel-tabset .panel-blue group="language"}
## R

**Visualise relationship**

```{r}
plot(gapminder_2007$lifeExp, gapminder_2007$gdpPercap)
```


```{r}
ggplot(gapminder_2007, aes(x = lifeExp, y = gdpPercap)) +
  geom_point()
```
:::

::: {.panel-tabset .panel-blue group="language"}
## R

**Log transform GDP**

```{r}
gapminder_2007$log_gdpPercap <- log(gapminder_2007$gdpPercap)
```
:::

::: {.panel-tabset .panel-blue group="language"}
## R

**Visualise relationship**

```{r}
plot(gapminder_2007$lifeExp, gapminder_2007$log_gdpPercap)
```
:::

## Model

::: {.panel-tabset .panel-blue group="language"}
## R

```{r}
model1 <- lm(lifeExp ~ log_gdpPercap, data = gapminder_2007)

summary(model1)
```
:::

## Assumption checking

::: {.panel-tabset .panel-blue group="language"}
## R

```{r}
plot(model1)
```
:::

Q-Q plot doesn't look great!


## Scatter plot with regression line

::: {.panel-tabset .panel-blue group="language"}
## Rs

```{r}
ggplot(gapminder_2007, aes(x = log_gdpPercap, y = lifeExp)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE)
```
:::


